---
title: "Exploration of Airline on-time arrival and delay data, 2003 ~ 2015"
author: "HanByul Yang"
date: "November 17, 2015"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
knitr::opts_chunk$set(comment = NA, echo=FALSE, message = FALSE, 
                      warnings = FALSE, cache=TRUE, cache.path='cache/',
                      fig.path='figure/')
# Loading required libraries
library(ggplot2)
library(ggthemes)
library(dplyr)
library(reshape2)
library(scales)
theme_set(theme_minimal(20))
```

```{r load_the_data}
# Load the data
df <- read.csv('263748736_92015_430_airline_delay_causes.csv')

# Remove unnecessary column "X" which indicates index.
drops <- c("X")
df <- df[, !(names(df) %in% drops)]

names(df)[names(df) == 'X.month'] <- 'month'
names(df)[names(df) == 'X.arr_delay'] <- 'arr_delay'
names(df)[names(df) == 'X.carrier_delay'] <- 'carrier_delay'
```

### Summary of the data set
```{r summary_of_data_set}
# Summarize the data set
dim(df)
names(df)
str(df)
summary(df)
```
204606 records dataset.

```{r data_wrangling}
df$date <- as.Date(paste(df$year, df$month, "01", sep="-"), format="%Y-%m-%d")
```

```{r make_new_dataFrame}
df_no_NAs <- df[complete.cases(df),]
df_sum <- df_no_NAs  %>%
  group_by(date, year) %>%
  summarise(
    flights = sum(arr_flights),
    delayed = sum(arr_del15),
    cancelled = sum(arr_cancelled),
    diverted = sum(arr_diverted)
  )
```

### numbers over year
```{r on_time_performance_over_year}
ggplot(aes(x = date, y = flights), data = df_sum) +
  geom_line(colour="red") +
  geom_line(aes(x = date, y=delayed), colour="green") +
  geom_line(aes(x = date, y=cancelled), colour="blue") +
  labs(title='',
       x='date', y='count')
```

red : flights, green : delayed, blue : cancelled.

```{r}
df_percent <- df_sum #data.frame(date=as.Date(character()), on_time=double())
df_percent$delayed <- df_sum$delayed / df_sum$flights
df_percent$cancelled <- df_sum$cancelled / df_sum$flights
df_percent$diverted <- df_sum$diverted / df_sum$flights
df_percent$on_time <- 1 - df_percent$delayed

drops <- c("flights", "year")
df_percent <- df_percent[, !(names(df_percent) %in% drops)]
df_p_long<-melt(df_percent, id.vars="date")
```

### on-time arrival performance over time (monthly)
```{r on_time_performance_monthly}
ggplot(df_p_long, aes(date, value, color=variable)) +
  geom_line()
```

```{r }
df_year <- df_no_NAs  %>%
  group_by(year) %>%
  summarise(
    flights = sum(arr_flights),
    delayed = sum(arr_del15),
    cancelled = sum(arr_cancelled),
    diverted = sum(arr_diverted)
  )
df_year <- df_year[complete.cases(df_year), ]
df_year_p <- df_year
df_year_p$delayed <- df_year$delayed / df_year$flights
df_year_p$cancelled <- df_year$cancelled / df_year$flights
df_year_p$diverted <- df_year$diverted / df_year$flights
df_year_p$on_time <- 1 - df_year_p$delayed

drops <- c("flights")
df_year_p <- df_year_p[, !(names(df_year_p) %in% drops)]
df_year_p_long<-melt(df_year_p, id.vars="year")
```

### on-time arrival performance over time (yearly)
```{r on_time_performance_yearly}
ggplot(df_year_p_long, aes(year,value,color=variable)) +
  geom_line() +
  # geom_point() +
  theme_economist() +
  scale_colour_economist()
```

### Flights by carrier
```{r carriers}
df_carrier_month_raw <- df %>%
  group_by(date, year, carrier_name) %>%
  summarise(
    flights = sum(arr_flights),
    delayed = sum(arr_del15),
    cancelled = sum(arr_cancelled),
    diverted = sum(arr_diverted)
  ) %>%
  transform(on_time = 1 - delayed/flights)

# remove NAs
df_carrier_month <- df_carrier_month_raw[complete.cases(df_carrier_month_raw),]

df_carriers <- df_carrier_month %>%
  group_by(carrier_name) %>%
  summarize(flights_avg = mean(flights),
            flights = sum(flights)) %>%
  arrange(desc(flights))

dim(df_carriers)
names(df_carriers)
str(df_carriers)
summary(df_carriers)
```
total 28 carriers.

### Cumulative relative flights
```{r Cumulative}
# top 5
cumflights = cumsum(df_carriers$flights)
cumrelflights = cumflights/ cumflights[28]
df_carriers$cumflights <- cumflights
df_carriers$cumrelflights <- cumrelflights
options(dplyr.width = Inf)
head(df_carriers)
top5_carriers <- df_carriers[1:5,]$carrier_name
df_top5_carrier <- subset(df_carrier_month, is.element(carrier_name, top5_carriers)) %>%
  group_by(year, carrier_name) %>%
  summarize(flights = sum(flights),
            delayed = sum(delayed),
            cancelled = sum(cancelled),
            diverted = sum(diverted)) %>%
  transform(on_time = 1 - delayed/flights)
```
top 5 covers over 50% of flights

### final data set
```{r final_dataset}
drops <- c("flights", "delayed", "cancelled", "diverted")
df_final <- df_top5_carrier[, !(names(df_top5_carrier) %in% drops)]

df_final2 <- dcast(df_final, year ~ carrier_name)
df_final2
```

### Top5 airlines on-time arrival performance over year
```{r top5_airlines_performance, fig.width=8.5, fig.height=5}
ggplot(df_final, aes(year, on_time, color=carrier_name)) +
  geom_line() +
  geom_point() +
  theme_economist() +
  guides(color=guide_legend(nrow=2))
```

```{r export_to_csv}
write.csv(df_final, file="airline_ontime_performace.csv", row.names=FALSE)
```

## References
http://stackoverflow.com/questions/20444970/how-to-make-a-date-column-by-year-in-r
http://stackoverflow.com/questions/4862178/remove-rows-with-nas-in-data-frame
http://stackoverflow.com/questions/15795071/how-to-add-legend-to-ggplot2-line-chart
http://www.statmethods.net/management/merging.html
http://stackoverflow.com/questions/27130610/legend-on-bottom-two-rows-wrapped-in-ggplot2-in-r
